---
title: "BIOS 611 — Programming in R"
output: html_document
---

This single file contains the answers and runnable code for Problems 1–8.

## Problem 1

`<-` assigns in the current evaluation environment. `<<-` searches enclosing
parent environments for an existing binding and updates the first one it finds;
if it finds none, it creates the binding in the global environment. This makes
`<<-` useful when a returned function (a closure) needs to update state held by
the function that created it.

## Problem 2

No new evaluation rule is strictly necessary. `1:10` is the infix form of the
primitive colon operator, equivalent to calling ``:`(1, 10)``. The existing
rule for evaluating a function/operator call explains why it returns the
integer vector `c(1, 2, ..., 10)`. A separate rule for `:` can be a convenient
shortcut, but it is not a fundamentally new kind of evaluation.

## Problem 3

The loop finishes with `test_value` equal to 100. The body of a `for` loop is
evaluated in the current evaluation environment; R does not create a fresh
lexical environment for every iteration. Therefore `<-` changes the binding in
the function environment if the loop is inside a function, or in the global
environment if it is run at the top level.

## Problem 4

The following test shows that assignments made in a `while` body remain
available after the loop:

```{r}
while_scope_test <- function() {
  test_value <- 10
  x <- 1
  while (x <= 100) {
    test_value <- x
    x <- x + 1
  }
  c(test_value = test_value, next_x = x)
}

while_scope_test()
```

The result is `test_value = 100` and `next_x = 101`, so `while` follows the
same environment rule as `for`: it does not make a new environment per
iteration.

## Problem 5

The returned function is a closure. It keeps two private bindings and updates
them with `<<-` on every call:

```{r}
make_fib_counter <- function() {
  first <- 1
  second <- 1

  function() {
    result <- first
    first <<- second
    second <<- result + second
    result
  }
}

fc <- make_fib_counter()
vapply(seq_len(6), function(...) fc(), numeric(1))
```

This produces `1, 1, 2, 3, 5, 8, ...` on successive calls.

## Problem 6

The same pattern can accept any initial pair:

```{r}
make_nocci <- function(c1, c2) {
  function() {
    result <- c1
    c1 <<- c2
    c2 <<- result + c2
    result
  }
}

nc <- make_nocci(2, 3)
vapply(seq_len(5), function(...) nc(), numeric(1))
```

For this example the values are `2, 3, 5, 8, 13, ...`. Every call to
`make_nocci` creates a separate enclosing environment, so separate counters do
not share state.

## Problem 7

`make_nocci` first creates `c1` and `c2` in its function execution environment,
then returns the inner function. The inner function retains a reference to that
environment after `make_nocci` returns; this retained function plus its
enclosing environment is a closure. On each invocation, `result <- c1` reads
the current first value. The two `<<-` assignments shift the pair forward: the
old second value becomes the new first value, and the sum becomes the new second
value. The returned `result` is therefore the next value in the sequence.

## Problem 8

Examples of R environments include:

* `.GlobalEnv`, the interactive workspace
* `baseenv()`, the base package environment
* `emptyenv()`, the parentless empty environment
* a function's execution environment/call frame
* the enclosing environment captured by a closure
* `new.env()`, a newly created user environment
* package environments such as `as.environment("package:stats")`
* namespace environments such as `asNamespace("stats")`
* parent environments obtained with `parent.env()`
* the environment returned by `environment()` inside a function

An environment maps names to bindings, has a parent environment, and can be
mutated. It is not the same thing as a list or data frame, even though all
three can store named values.

## Verification

The following checks exercise the functions above:

```{r}
stopifnot(identical(unname(vapply(seq_len(6), function(...) fc(), numeric(1))),
                    c(1, 1, 2, 3, 5, 8)))
stopifnot(identical(unname(vapply(seq_len(5), function(...) nc(), numeric(1))),
                    c(2, 3, 5, 8, 13)))
stopifnot(identical(unname(while_scope_test()), c(100, 101)))
```
